---
title: "Codes for Dissertation: Hawkes Processes Modelling of the COVID-19 Pandemic in Israel Using Whittle Estimation Method"
author: "Jia Xin Lin"
output: pdf_document
---
# Whole Israel
## Data cleaning and manipulation
```{r setup_data_cleaning, include=FALSE, message=FALSE, warning=FALSE, echo=FALSE}
#Packages used
library(rio)
library(dplyr)
library(ggplot2)
library(grid)
library(hawkesbow)
library(foreach)
library(doParallel)
library(SuperGauss)

#Source from Cheysson's GitHub: https://github.com/fcheysson/code-spectral-hawkes/tree/master/case%20study
source("whittle_periodogram.R")

#Importing data
geo = import("geographic_data_copy.csv", setclass = "tibble")
#Filtering out the twon names in Hebrew
geo2 <- geo[-3]


#Checking NA-s
sapply(geo2, function(x) sum(is.na(x)))
# 305240 NA-s in agas_code

#Summary about the data set
summary(geo2)
```

```{r creating_variables, message=FALSE, warning=FALSE, echo=FALSE}
#Creating variables for Cases: new_cases_per_day AND seven_day_avg_case
geo_cases <- select(geo2, town_code, date, accumulated_cases, new_cases_on_date)
geo_cases$accumulated_cases[geo_cases$accumulated_cases == '<15'] <- 15
geo_cases$accumulated_cases <- as.numeric(geo_cases$accumulated_cases)
geo_cases_date <- geo_cases %>% group_by(date) %>% 
  summarise(sum_acc_cases = sum(accumulated_cases),
            .groups = 'drop')
geo_cases_date$new_cases_per_day <- geo_cases_date$sum_acc_cases
for (i in 2:length(geo_cases_date$sum_acc_cases)) {
  geo_cases_date$new_cases_per_day[i] <- geo_cases_date$sum_acc_cases[i]-geo_cases_date$sum_acc_cases[i-1]
}
geo_cases_date$seven_day_avg_case <- geo_cases_date$new_cases_per_day
for (i in 7:length(geo_cases_date$new_cases_per_day)) {
  geo_cases_date$seven_day_avg_case[i] <- (geo_cases_date$new_cases_per_day[i]+geo_cases_date$new_cases_per_day[i-1]+geo_cases_date$new_cases_per_day[i-2]+geo_cases_date$new_cases_per_day[i-3]+geo_cases_date$new_cases_per_day[i-4]+geo_cases_date$new_cases_per_day[i-5]+geo_cases_date$new_cases_per_day[i-6])/7
}

#Creating variables for Mortality: new_deaths_per_day AND seven_day_avg_death
geo_deaths <- select(geo2, town_code, date, accumulated_deaths, new_deaths_on_date)
geo_deaths$accumulated_deaths[geo_deaths$accumulated_deaths == '<15'] <- 15
geo_deaths$accumulated_deaths <- as.numeric(geo_deaths$accumulated_deaths)
geo_deaths_date <- geo_deaths %>% group_by(date) %>% 
  summarise(sum_acc_deaths = sum(accumulated_deaths),
            .groups = 'drop')
geo_deaths_date$new_deaths_per_day <- geo_deaths_date$sum_acc_deaths
for (i in 2:length(geo_deaths_date$sum_acc_deaths)) {
  geo_deaths_date$new_deaths_per_day[i] <- geo_deaths_date$sum_acc_deaths[i]-geo_deaths_date$sum_acc_deaths[i-1]
}
geo_deaths_date$seven_day_avg_death <- geo_deaths_date$new_deaths_per_day
for (i in 7:length(geo_deaths_date$new_deaths_per_day)) {
  geo_deaths_date$seven_day_avg_death[i] <- (geo_deaths_date$new_deaths_per_day[i]+geo_deaths_date$new_deaths_per_day[i-1]+geo_deaths_date$new_deaths_per_day[i-2]+geo_deaths_date$new_deaths_per_day[i-3]+geo_deaths_date$new_deaths_per_day[i-4]+geo_deaths_date$new_deaths_per_day[i-5]+geo_deaths_date$new_deaths_per_day[i-6])/7
}
```

## Making Plots: daily_cases_deaths_Israel
```{r plots_whole, message=FALSE, warning=FALSE, echo=FALSE, fig.width=18, fig.height=6}
#All Cases
NewCase_Date <- geo_cases_date %>%
        ggplot(aes(x = date)) + 
        geom_bar(aes(y = new_cases_per_day, fill = "Number of cases"),stat = "identity") +
        xlab("Date") + ylab("Daily Cases") +
        ggtitle("Cases in Israel (11 March 2020 - 28 May 2023)") +
        theme(plot.title = element_text(lineheight=.8, face="bold"))  +
        geom_line(aes(y=seven_day_avg_case, colour = "Cases (7-day average)"), size = 1.2, alpha = 0.85) +
        scale_fill_manual(name = "", values = c("Number of cases" = "orange2")) +
        scale_colour_manual(name = "", values = c("Cases (7-day average)" = "sienna")) +
        scale_x_date(date_breaks = "3 month",
                     date_minor_breaks = "1 month", date_labels = "%b-%y")

#All Deaths
Death_Date <- geo_deaths_date %>%
            ggplot(aes(x = date))+
            geom_bar(aes(y = new_deaths_per_day ,fill = "Number of Deaths"),stat = "identity", position = "stack") +
            geom_line(aes(y=seven_day_avg_death, colour = "Deaths (7-day average)"), size = 1.2, alpha = 0.9) +
            xlab("Date")+ylab("Daily Deaths") +
            ggtitle("Deaths in Israel (11 March 2020 - 28 May 2023)") +
            theme(plot.title = element_text(lineheight=.8, face="bold"))+
            scale_fill_manual(name = "", values = c("Number of Deaths" = "firebrick3"))+
            scale_colour_manual(name = "", values = c("Deaths (7-day average)" = "darkred"))+
            scale_x_date(date_breaks = "3 month",
                         date_minor_breaks = "1 month", date_labels = "%b-%y")

#Saving as png
grid.newpage()
Case_Death <- arrangeGrob(cbind(ggplotGrob(NewCase_Date + theme(legend.position="none")), ggplotGrob(Death_Date+ theme(legend.position="none")), size = "last"))
grid.draw(Case_Death)
ggsave("Case_Death.png", Case_Death, width = 50, height = 17, units = "cm")


#First Wave: Feb - May 2020
#Cases
geo_cases_date1 <- geo_cases_date %>%
                   filter(date < '2020-5-16')
NewCase_Date1 <- geo_cases_date1 %>%
        ggplot(aes(x = date)) +
        geom_bar(aes(y = new_cases_per_day, fill = "Number of cases"),stat = "identity") +
        xlab("Date") + ylab("Daily Cases") +
        ggtitle("Cases in Israel (Feb - May 2020)") +
        theme(plot.title = element_text(lineheight=.8, face="bold"))  +
        geom_line(aes(y=seven_day_avg_case, colour = "Cases (7-day average)"), size = 1.2, alpha = 0.85) +
        scale_fill_manual(name = "", values = c("Number of cases" = "orange2")) +
        scale_colour_manual(name = "", values = c("Cases (7-day average)" = "sienna")) +
        scale_x_date(date_breaks = "1 week",
                     date_minor_breaks = "1 week", date_labels = "%d-%b")

#Mortality
geo_deaths_date1 <- geo_deaths_date %>%
                   filter(date < '2020-5-16')
Death_Date1 <- geo_deaths_date1 %>%
            ggplot(aes(x = date))+
            geom_bar(aes(y = new_deaths_per_day ,fill = "Number of Deaths"),stat = "identity", position = "stack") +
            geom_line(aes(y=seven_day_avg_death, colour = "Deaths (7-day average)"), size = 1.2, alpha = 0.9) +
            xlab("Date")+ylab("Daily Deaths") +
            ggtitle("Deaths in Israel (Feb - May 2020)") +
            theme(plot.title = element_text(lineheight=.8, face="bold"))+
            scale_fill_manual(name = "", values = c("Number of Deaths" = "firebrick3"))+
            scale_colour_manual(name = "", values = c("Deaths (7-day average)" = "darkred"))+
            scale_x_date(date_breaks = "1 week",
                         date_minor_breaks = "1 week", date_labels = "%d-%b")

grid.newpage()
Case_Death1 <- arrangeGrob(cbind(ggplotGrob(NewCase_Date1 + theme(legend.position="none")), ggplotGrob(Death_Date1+ theme(legend.position="none")), size = "last"))
grid.draw(Case_Death1)
ggsave("Case_Death1.png", Case_Death, width = 50, height = 17, units = "cm")

#Similarly for other waves
```

## Fitting Model
```{r fitting_whole, message=FALSE, warning=FALSE, echo=FALSE}
##########################Exponential for 1st wave
#Infection modelling
whittle(geo_cases_date1$new_cases_per_day, "Exponential", 1)$opt$par
#Mortality modelling
whittle(geo_deaths_date1$new_deaths_per_day, "Exponential", 1)$opt$par
#Repeat for other waves

##########################Power-law for 1st wave
#Infection modelling
whittle(geo_cases_date1$new_cases_per_day, "PowerLaw", 1)$opt$par
#Mortality modelling
whittle(geo_deaths_date1$new_deaths_per_day, "PowerLaw", 1)$opt$par
#Repeat for other waves

#########################Gaussian for 1st wave
#Infection modelling
gau_cases1 = NULL
value1 = Inf
for (i in 1:100) {
  temp = whittle(geo_cases_date1$new_cases_per_day, "Gaussian", 1)
  if (temp$opt$value < value1) {
      gau_cases1 = temp
      value1 = temp$opt$value
  }
}
gau_cases1$par #results: 163.46304811   0.97701181   0.01000221   4.12891904

#Mortality modelling
gau_deaths1 = NULL
value1 = Inf
for (i in 1:100) {
  temp = whittle(geo_deaths_date1$new_deaths_per_day, "Gaussian", 1)
  if (temp$opt$value < value1) {
      gau_deaths1 = temp
      value1 = temp$opt$value
  }
}
gau_deaths1$par #results: 14.6172738  0.9759815  1.9035884  0.5299262
#due to the randomness in Gaussian kernel, the code might give slightly different value when running the codes again
#Repeat for other waves 
```

## Goodness of fit for Gaussian kernel
```{r goodness_whole_gau, message=FALSE, warning=FALSE, echo=FALSE}
# This goodness of fit test is inspired based on Cheysson's code from: https://github.com/fcheysson/code-spectral-hawkes/tree/master/case%20study

#Goodness of fit for 1st wave
init = gau_cases1$par
y = geo_cases_date1$new_cases_per_day
n = length(y)
Nsup = ceiling((n-1)/2)
Ninf = - floor((n-1)/2)
omega = 2 * pi * Ninf:Nsup / n

kern = function(x) ifelse(abs(x) <= pi, 1.5 * (1.0 - (x / pi)^2), 0.0)
kernh = function(x, h) kern(x / h)
h = 0.10

k = kernh(2* pi * 0:(n-1) / n, h)
q = function(J) 1 / (n*h) * SuperGauss::toep.mult(k, J)

I = Mod(fft(y - mean(y)))^2 / n
I = c(I[(Nsup+2):n], I[1:(Nsup+1)])
f = gau_cases1$kernel$f1(omega, trunc = 10L)
S = 2*pi*sqrt(h) * sum( q(I/f - 1)^2 )

n.sim = 1e3
cl = makePSOCKcluster(3) # use multicore, set to the number of our cores
clusterExport(cl, c("n", "Nsup", "omega", "h", "k", "f", "q", "init"))
registerDoParallel(cl)
Ss = foreach (i = icount(n.sim), .combine = c, .multicombine = TRUE, .inorder = FALSE,
              .packages = c("SuperGauss", "hawkesbow")) %dopar% {
                Iboots = f * rexp(n)
                Iboots[n - Nsup] = f[n - Nsup] * rchisq(1, df = 1)
                model = whittle_periodogram(c(Iboots[(n-Nsup+1):n], Iboots[1:(n-Nsup-1)]), 
                                            kern = "gauss", binsize = 1, init = init)
                fboots = model$f1(omega, trunc = 10L)
                return( 2*pi*sqrt(h) * sum( q(Iboots/fboots - 1)^2 ) )
              }
stopCluster(cl)

muh = 1/sqrt(h) * 12*pi/5
tau2 = 2672*pi^2/385
 plot(Ss)
plot(z<-0:100, dnorm(z, mean = muh, sd = sqrt(tau2)), col=2, type="l")
lines(density(Ss))
abline(v=S, col="grey")
cat("mean(Ss) = ", mean(Ss), " and     sd(Ss) = ", sd(Ss),
    "\n     muh = ", muh,      " and sqrt(tau2) = ", sqrt(tau2), 
    sep = "")
c(muh - 1.96*sqrt(tau2), muh + 1.96*sqrt(tau2))

#results:
# > cat("mean(Ss) = ", mean(Ss), " and     sd(Ss) = ", sd(Ss),
# +     "\n     muh = ", muh,      " and sqrt(tau2) = ", sqrt(tau2), 
# +     sep = "")
# mean(Ss) = 33.53907 and     sd(Ss) = 71.06017
#      muh = 23.84301 and sqrt(tau2) = 8.276329
# > c(muh - 1.96*sqrt(tau2), muh + 1.96*sqrt(tau2))
# [1]  7.621407 40.064616

#Repeat for other waves
```


# Dividing Israel into Areas: producing 'geo_area_loc.csv' file
The below code is a part of creating 'geo_area_loc.csv' file, after this file was created I used Excel and 'geo_population.csv' to create the 'geo_area_loc_std.csv' file, which includes the standardised cases.
```{r splti_into_areas, message=FALSE, warning=FALSE, echo=FALSE}
#################################North Israel
#Cases data
geo_cases <- select(geo2, town_code, date, accumulated_cases, new_cases_on_date)
geo_cases$accumulated_cases[geo_cases$accumulated_cases == '<15'] <- 15
geo_cases$accumulated_cases <- as.numeric(geo_cases$accumulated_cases)

######## Qiryat Shemona
geo_cases_2800 <- geo_cases %>% filter(geo_cases$town_code == 2800) %>% 
                            group_by(date) %>% 
                            summarise(accumulated_cases = sum(accumulated_cases),.groups = 'drop')

geo_cases_2800$new_cases_per_day <- geo_cases_2800$accumulated_cases
for (i in 2:length(geo_cases_2800$accumulated_cases)) {
  geo_cases_2800$new_cases_per_day[i] <- geo_cases_2800$accumulated_cases[i]-geo_cases_2800$accumulated_cases[i-1]
}
geo_cases_2800$seven_day_avg_case <- geo_cases_2800$new_cases_per_day
for (i in 7:length(geo_cases_2800$new_cases_per_day)) {
  geo_cases_2800$seven_day_avg_case[i] <- (geo_cases_2800$new_cases_per_day[i]+geo_cases_2800$new_cases_per_day[i-1]+geo_cases_2800$new_cases_per_day[i-2]+geo_cases_2800$new_cases_per_day[i-3]+geo_cases_2800$new_cases_per_day[i-4]+geo_cases_2800$new_cases_per_day[i-5]+geo_cases_2800$new_cases_per_day[i-6])/7
}

#Creating geo_area_loc data frame
geo_area_loc <- select(geo_cases_2800, date)
#Filling up dataframe
geo_area_loc$area <- rep("North", 1174)
geo_area_loc$town <- rep("Qiryat Shemona", 1174)
geo_area_loc$daily_cases <- geo_cases_2800$new_cases_per_day
geo_area_loc$avg_cases <- geo_cases_2800$seven_day_avg_case


######## Afula
geo_cases_7700 <- geo_cases %>% filter(geo_cases$town_code == 7700) %>% 
                            group_by(date) %>% 
                            summarise(accumulated_cases = sum(accumulated_cases),.groups = 'drop')

geo_cases_7700$new_cases_per_day <- geo_cases_7700$accumulated_cases
for (i in 2:length(geo_cases_7700$accumulated_cases)) {
  geo_cases_7700$new_cases_per_day[i] <- geo_cases_7700$accumulated_cases[i]-geo_cases_7700$accumulated_cases[i-1]
}
geo_cases_7700$seven_day_avg_case <- geo_cases_7700$new_cases_per_day
for (i in 7:length(geo_cases_7700$new_cases_per_day)) {
  geo_cases_7700$seven_day_avg_case[i] <- (geo_cases_7700$new_cases_per_day[i]+geo_cases_7700$new_cases_per_day[i-1]+geo_cases_7700$new_cases_per_day[i-2]+geo_cases_7700$new_cases_per_day[i-3]+geo_cases_7700$new_cases_per_day[i-4]+geo_cases_7700$new_cases_per_day[i-5]+geo_cases_7700$new_cases_per_day[i-6])/7
}
#Summarise the data
data_7700 <- data.frame(date = geo_cases_7700$date,
                        area = rep("North",1174),
                        town = rep("Afula",1174),
                        daily_cases = geo_cases_7700$new_cases_per_day,
                        avg_cases = geo_cases_7700$seven_day_avg_case)
#Filling up dataframe
geo_area_loc<- rbind(geo_area_loc, data_7700)


####### Karmi'el
geo_cases_1139 <- geo_cases %>% filter(geo_cases$town_code == 1139) %>% 
                            group_by(date) %>% 
                            summarise(accumulated_cases = sum(accumulated_cases),.groups = 'drop')

geo_cases_1139$new_cases_per_day <- geo_cases_1139$accumulated_cases
for (i in 2:length(geo_cases_1139$accumulated_cases)) {
  geo_cases_1139$new_cases_per_day[i] <- geo_cases_1139$accumulated_cases[i]-geo_cases_1139$accumulated_cases[i-1]
}
geo_cases_1139$seven_day_avg_case <- geo_cases_1139$new_cases_per_day
for (i in 7:length(geo_cases_1139$new_cases_per_day)) {
  geo_cases_1139$seven_day_avg_case[i] <- (geo_cases_1139$new_cases_per_day[i]+geo_cases_1139$new_cases_per_day[i-1]+geo_cases_1139$new_cases_per_day[i-2]+geo_cases_1139$new_cases_per_day[i-3]+geo_cases_1139$new_cases_per_day[i-4]+geo_cases_1139$new_cases_per_day[i-5]+geo_cases_1139$new_cases_per_day[i-6])/7
}

data_1139 <- data.frame(date = geo_cases_1139$date,
                        area = rep("North",1174),
                        town = rep("Karmi'el",1174),
                        daily_cases = geo_cases_1139$new_cases_per_day,
                        avg_cases = geo_cases_1139$seven_day_avg_case)

geo_area_loc<- rbind(geo_area_loc, data_1139)


######## Haifa
geo_cases_4000 <- geo_cases %>% filter(geo_cases$town_code == 4000) %>% 
                            group_by(date) %>% 
                            summarise(accumulated_cases = sum(accumulated_cases),.groups = 'drop')

geo_cases_4000$new_cases_per_day <- geo_cases_4000$accumulated_cases
for (i in 2:length(geo_cases_4000$accumulated_cases)) {
  geo_cases_4000$new_cases_per_day[i] <- geo_cases_4000$accumulated_cases[i]-geo_cases_4000$accumulated_cases[i-1]
}
geo_cases_4000$seven_day_avg_case <- geo_cases_4000$new_cases_per_day
for (i in 7:length(geo_cases_4000$new_cases_per_day)) {
  geo_cases_4000$seven_day_avg_case[i] <- (geo_cases_4000$new_cases_per_day[i]+geo_cases_4000$new_cases_per_day[i-1]+geo_cases_4000$new_cases_per_day[i-2]+geo_cases_4000$new_cases_per_day[i-3]+geo_cases_4000$new_cases_per_day[i-4]+geo_cases_4000$new_cases_per_day[i-5]+geo_cases_4000$new_cases_per_day[i-6])/7
}

data_4000 <- data.frame(date = geo_cases_4000$date,
                        area = rep("North",1174),
                        town = rep("Haifa",1174),
                        daily_cases = geo_cases_4000$new_cases_per_day,
                        avg_cases = geo_cases_4000$seven_day_avg_case)

geo_area_loc<- rbind(geo_area_loc, data_4000)

#Similarly extracting data for other areas, then add them into geo_area_loc
```

To produce the plots presented in the thesis, we can use the 'geo_area_loc_std.csv' file directly.
```{r load_area_loc, message=FALSE, warning=FALSE, echo=FALSE}
geo_area_loc_std = import("geo_area_loc_std.csv", setclass = "tibble")
geo_area_loc_std$date <- as.Date(geo_area_loc_std$date)
```

```{r plots_north, message=FALSE, warning=FALSE, echo=FALSE, fig.width=18, fig.height=6}
######## North Israel over the whole pandemic
#Seven-day average cases plots for North Israel
Case_Date_North <- geo_area_loc_std %>%
                      filter(area=="North") %>%
        ggplot(aes(x = date)) + 
        xlab("Date") + ylab("Daily Cases") +
        ggtitle("Cases in Northern Israel \n (11 March 2020 - 28 May 2023)") +
        theme(plot.title = element_text(lineheight=.8, face="bold"))  +
        geom_line(aes(y=avg_cases, colour = town)) +
        scale_x_date(date_breaks = "3 month",
                     date_minor_breaks = "1 month", date_labels = "%b-%y")

#Standardised average cases plot for North Israel
Case_Date_North_std <- geo_area_loc_std %>%
                      filter(area=="North") %>%
        ggplot(aes(x = date)) + 
        xlab("Date") + ylab("Standardised Daily Cases") +
        ggtitle("Standardised Cases (per 100000 people) in Northern Israel \n (11 March 2020 - 28 May 2023)") +
        theme(plot.title = element_text(lineheight=.8, face="bold"))  +
        geom_line(aes(y=std_avg_cases, colour = town)) +
        scale_x_date(date_breaks = "3 month",
                     date_minor_breaks = "1 month", date_labels = "%b-%y")
grid.newpage()
Case_Date_North_with_std <- arrangeGrob(cbind(ggplotGrob(Case_Date_North+ theme(legend.position="none")), ggplotGrob(Case_Date_North_std), size = "last"))
grid.draw(Case_Date_North_with_std)
ggsave("Case_Date_North_with_std.png", Case_Date_North_with_std, width = 50, height = 17, units = "cm")


########Plots for 1-4th wave North Israel without Big city (Haifa)
#Seven-day average cases plots
Case_Date_North_Loc1 <- geo_area_loc_std_old %>%
                      filter(area=="North") %>%
                      filter(date <= '2021-11-15') %>%
        ggplot(aes(x = date)) + 
        xlab("Date") + ylab("Daily Cases") +
        ggtitle("Cases in Israel by Three Northern Towns \n (11 March 2020 - 15 Nov 2021)") +
        theme(plot.title = element_text(lineheight=.8, face="bold"))  +
        geom_line(aes(y=avg_cases, colour = town)) +
        scale_x_date(date_breaks = "3 month",
                     date_minor_breaks = "1 month", date_labels = "%b-%y")

#Standardised average cases plot
Case_Date_North_Loc_std1 <- geo_area_loc_std_old %>%
                      filter(area=="North") %>%
                      filter(date <= '2021-11-15') %>%
        ggplot(aes(x = date)) + 
        xlab("Date") + ylab("Standardised Daily Cases") +
        ggtitle("Standardised Cases (per 100000 people) in Israel by Three Northern Towns \n (11 March 2020 - 15 Nov 2021)") +
        theme(plot.title = element_text(lineheight=.8, face="bold"))  +
        geom_line(aes(y=std_avg_cases, colour = town)) +
        scale_x_date(date_breaks = "3 month",
                     date_minor_breaks = "1 month", date_labels = "%b-%y")
grid.newpage()
Case_Date_North_Loc_with_std1 <- arrangeGrob(cbind(ggplotGrob(Case_Date_North_Loc1+ theme(legend.position="none")), ggplotGrob(Case_Date_North_Loc_std1), size = "last"))
grid.draw(Case_Date_North_Loc_with_std1)
ggsave("Case_Date_North_Loc_with_std1.png", Case_Date_North_Loc_with_std1, width = 50, height = 17, units = "cm")


########Plots since 5th wave North Israel without Big city (Haifa)
#Seven-day average cases plots
Case_Date_North_Loc2 <- geo_area_loc_std_old %>%
                      filter(area=="North") %>%
                      filter(date > '2021-11-15') %>%
        ggplot(aes(x = date)) + 
        xlab("Date") + ylab("Daily Cases") +
        ggtitle("Cases in Israel by Three Northern Towns \n (16 Nov 2021 - 28 May 2023)") +
        theme(plot.title = element_text(lineheight=.8, face="bold"))  +
        geom_line(aes(y=avg_cases, colour = town)) +
        scale_x_date(date_breaks = "3 month",
                     date_minor_breaks = "1 month", date_labels = "%b-%y")

#Standardised average cases plot
Case_Date_North_Loc_std2 <- geo_area_loc_std_old %>%
                      filter(area=="North") %>%
                      filter(date > '2021-11-15') %>%
        ggplot(aes(x = date)) + 
        xlab("Date") + ylab("Standardised Daily Cases") +
        ggtitle("Standardised Cases (per 100000 people) in Israel by Three Northern Towns \n (16 Nov 2021 - 28 May 2023)") +
        theme(plot.title = element_text(lineheight=.8, face="bold"))  +
        geom_line(aes(y=std_avg_cases, colour = town)) +
        scale_x_date(date_breaks = "3 month",
                     date_minor_breaks = "1 month", date_labels = "%b-%y")
grid.newpage()
Case_Date_North_Loc_with_std2 <- arrangeGrob(cbind(ggplotGrob(Case_Date_North_Loc2+ theme(legend.position="none")), ggplotGrob(Case_Date_North_Loc_std2), size = "last"))
grid.draw(Case_Date_North_Loc_with_std2)
ggsave("Case_Date_North_Loc_with_std2.png", Case_Date_North_Loc_with_std2, width = 50, height = 17, units = "cm")

#Similarly for other areas
```

# Selected Five Cities
```{r plots_five, message=FALSE, warning=FALSE, echo=FALSE, fig.width=18, fig.height=6}
######## Five Cities over the whole pandemic
#Seven-day average cases plots for the selected Five Cities
Case_Date_Location <- geo_area_loc_std %>%
                      filter(town=="Haifa" | town=="Tel Aviv-Yafo" | town=="Jerusalem" | town=="Be'er Sheva" | town=="Lod") %>%
        ggplot(aes(x = date)) + 
        xlab("Date") + ylab("Daily Cases") +
        ggtitle("Cases in Israel by Five Major Cities \n (11 March 2020 - 28 May 2023)") +
        theme(plot.title = element_text(lineheight=.8, face="bold"))  +
        geom_line(aes(y=avg_cases, colour = town)) +
        scale_x_date(date_breaks = "3 month",
                     date_minor_breaks = "1 month", date_labels = "%b-%y")

#Standardised average cases plot for the selected Five Cities
Case_Date_Location_std <- geo_area_loc_std %>%
                      filter(town=="Haifa" | town=="Tel Aviv-Yafo" | town=="Jerusalem" | town=="Be'er Sheva" | town=="Lod") %>%
        ggplot(aes(x = date)) + 
        xlab("Date") + ylab("Standardised Daily Cases") +
        ggtitle("Standardised Cases (per 100000 people) in Israel by Five Major Cities \n (11 March 2020 - 28 May 2023)") +
        theme(plot.title = element_text(lineheight=.8, face="bold"))  +
        geom_line(aes(y=std_avg_cases, colour = town)) +
        scale_x_date(date_breaks = "3 month",
                     date_minor_breaks = "1 month", date_labels = "%b-%y")
grid.newpage()
Case_Date_Location_with_std <- arrangeGrob(cbind(ggplotGrob(Case_Date_Location+ theme(legend.position="none")), ggplotGrob(Case_Date_Location_std), size = "last"))
grid.draw(Case_Date_Location_with_std)
ggsave("Case_Date_Location_with_std.png", Case_Date_Location_with_std, width = 50, height = 17, units = "cm")


########Plots for 1-4th wave Five Cities
#Seven-day average cases plots
Case_Date_Location1 <- geo_area_loc_std %>%
                      filter(town=="Haifa" | town=="Tel Aviv-Yafo" | town=="Jerusalem" | town=="Be'er Sheva" | town=="Lod") %>%
                      filter(date <= '2021-11-15') %>%
        ggplot(aes(x = date)) + 
        xlab("Date") + ylab("Daily Cases") +
        ggtitle("Cases in Israel by Five Major Cities \n (11 March 2020 - 15 Nov 2021)") +
        theme(plot.title = element_text(lineheight=.8, face="bold"))  +
        geom_line(aes(y=avg_cases, colour = town)) +
        scale_x_date(date_breaks = "3 month",
                     date_minor_breaks = "1 month", date_labels = "%b-%y")

#Standardised average cases plot
Case_Date_Location_std1 <- geo_area_loc_std %>%
                      filter(town=="Haifa" | town=="Tel Aviv-Yafo" | town=="Jerusalem" | town=="Be'er Sheva" | town=="Lod") %>%
                      filter(date <= '2021-11-15') %>%
        ggplot(aes(x = date)) + 
        xlab("Date") + ylab("Standardised Daily Cases") +
        ggtitle("Standardised Cases (per 100000 people) in Israel by Five Major Cities \n (11 March 2020 - 15 Nov 2021)") +
        theme(plot.title = element_text(lineheight=.8, face="bold"))  +
        geom_line(aes(y=std_avg_cases, colour = town)) +
        scale_x_date(date_breaks = "3 month",
                     date_minor_breaks = "1 month", date_labels = "%b-%y")
grid.newpage()
Case_Date_Location_with_std1 <- arrangeGrob(cbind(ggplotGrob(Case_Date_Location1+ theme(legend.position="none")), ggplotGrob(Case_Date_Location_std1), size = "last"))
grid.draw(Case_Date_Location_with_std1)
ggsave("Case_Date_Location_with_std1.png", Case_Date_Location_with_std1, width = 50, height = 17, units = "cm")


########Plots since 5th wave Five Cities
#Seven-day average cases plots
Case_Date_Location2 <- geo_area_loc_std %>%
                      filter(town=="Haifa" | town=="Tel Aviv-Yafo" | town=="Jerusalem" | town=="Be'er Sheva" | town=="Lod") %>%
                      filter(date > '2021-11-15') %>%
        ggplot(aes(x = date)) + 
        xlab("Date") + ylab("Daily Cases") +
        ggtitle("Cases in Israel by Five Major Cities \n (16 Nov 2021 - 28 May 2023)") +
        theme(plot.title = element_text(lineheight=.8, face="bold"))  +
        geom_line(aes(y=avg_cases, colour = town)) +
        scale_x_date(date_breaks = "3 month",
                     date_minor_breaks = "1 month", date_labels = "%b-%y")

#Standardised average cases plot
Case_Date_Location_std2 <- geo_area_loc_std %>%
                      filter(town=="Haifa" | town=="Tel Aviv-Yafo" | town=="Jerusalem" | town=="Be'er Sheva" | town=="Lod") %>%
                      filter(date > '2021-11-15') %>%
        ggplot(aes(x = date)) + 
        xlab("Date") + ylab("Standardised Daily Cases") +
        ggtitle("Standardised Cases (per 100000 people) in Israel by Five Major Cities \n (16 Nov 2021 - 28 May 2023)") +
        theme(plot.title = element_text(lineheight=.8, face="bold"))  +
        geom_line(aes(y=std_avg_cases, colour = town)) +
        scale_x_date(date_breaks = "3 month",
                     date_minor_breaks = "1 month", date_labels = "%b-%y")
grid.newpage()
Case_Date_Location_with_std2 <- arrangeGrob(cbind(ggplotGrob(Case_Date_Location2+ theme(legend.position="none")), ggplotGrob(Case_Date_Location_std2), size = "last"))
grid.draw(Case_Date_Location_with_std2)
ggsave("Case_Date_Location_with_std2.png", Case_Date_Location_with_std2, width = 50, height = 17, units = "cm")
```

## Fitting the five cities
```{r splitting_five, message=FALSE, warning=FALSE, echo=FALSE}
geo_five_cities <- geo_area_loc_std %>%
                  filter(town=="Haifa" | town=="Tel Aviv-Yafo" | town=="Jerusalem" | town=="Be'er Sheva" | town=="Lod")

#To fit a model we also need to split them into 6 waves
geo_five_cities1 <- geo_five_cities %>%
                   filter(date < '2020-5-16')
geo_five_cities2 <- geo_five_cities %>%
                   filter(between(date, as.Date('2020-5-16'), as.Date('2020-11-15')))
geo_five_cities3 <- geo_five_cities %>%
                   filter(between(date, as.Date('2020-11-16'), as.Date('2021-4-15')))
geo_five_cities4 <- geo_five_cities %>%
                   filter(between(date, as.Date('2021-6-16'), as.Date('2021-11-15')))
geo_five_cities5 <- geo_five_cities %>%
                   filter(between(date, as.Date('2021-12-16'), as.Date('2022-4-30')))
geo_five_cities6 <- geo_five_cities %>%
                   filter(date > '2022-5-31')
```

```{r fitting_five, message=FALSE, warning=FALSE, echo=FALSE}
############################# Jerusalem
#Getting Jerusalem data
geo_five_cities1_3000 <- geo_five_cities1 %>%
                  filter(town=="Jerusalem")
geo_five_cities2_3000 <- geo_five_cities2 %>%
                  filter(town=="Jerusalem")
geo_five_cities3_3000 <- geo_five_cities3 %>%
                  filter(town=="Jerusalem")
geo_five_cities4_3000 <- geo_five_cities4 %>%
                  filter(town=="Jerusalem")
geo_five_cities5_3000 <- geo_five_cities5 %>%
                  filter(town=="Jerusalem")
geo_five_cities6_3000 <- geo_five_cities6 %>%
                  filter(town=="Jerusalem")


########Fitting 1st wave
gau_cases1_3000 = NULL
value1_3000 = Inf
for (i in 1:100) {
  temp = whittle(geo_five_cities1_3000$avg_cases, "Gaussian", 1)
  if (temp$opt$value < value1_3000) {
      gau_cases1_3000 = temp
      value1_3000 = temp$opt$value
  }
}
gau_cases1_3000$par #results: 1.65208022 0.99000000 1.53284253 0.05620799


########Fitting 2nd wave
gau_cases2_3000 = NULL
value2_3000 = Inf
for (i in 1:100) {
  temp = whittle(geo_five_cities2_3000$avg_cases, "Gaussian", 1)
  if (temp$opt$value < value2_3000) {
      gau_cases2_3000 = temp
      value2_3000 = temp$opt$value
  }
}
gau_cases2_3000$par #results: 0.4203269 0.9900000 0.1551308 0.5138064


########Fitting 3rd wave
gau_cases3_3000 = NULL
value3_3000 = Inf
for (i in 1:100) {
  temp = whittle(geo_five_cities3_3000$avg_cases, "Gaussian", 1)
  if (temp$opt$value < value3_3000) {
      gau_cases3_3000 = temp
      value3_3000 = temp$opt$value
  }
}
gau_cases3_3000$par #results: 0.9862177 0.9900000 0.0100000 1.4254454


########Fitting 4th wave
gau_cases4_3000 = NULL
value4_3000 = Inf
for (i in 1:100) {
  temp = whittle(geo_five_cities4_3000$avg_cases, "Gaussian", 1)
  if (temp$opt$value < value4_3000) {
      gau_cases4_3000 = temp
      value4_3000 = temp$opt$value
  }
}
gau_cases4_3000$par #results: 0.3950601 0.9900000 0.1768107 0.5795835


########Fitting 5th wave
gau_cases5_3000 = NULL
value5_3000 = Inf
for (i in 1:100) {
  temp = whittle(geo_five_cities5_3000$avg_cases, "Gaussian", 1)
  if (temp$opt$value < value5_3000) {
      gau_cases5_3000 = temp
      value5_3000 = temp$opt$value
  }
}
gau_cases5_3000$par #results: 11.6837766  0.9900000  0.0100000  0.8081723


########Fitting 6th wave
gau_cases6_3000 = NULL
value6_3000 = Inf
for (i in 1:100) {
  temp = whittle(geo_five_cities6_3000$avg_cases, "Gaussian", 1)
  if (temp$opt$value < value6_3000) {
      gau_cases6_3000 = temp
      value6_3000 = temp$opt$value
  }
}
gau_cases6_3000$par #results: 0.214170 0.990000 0.010000 2.484289

#due to the randomness in Gaussian kernel, the code might give slightly different value when running the codes again
#Repeat for other cities
```



